First run at causal models using daylight hours. Only done total so far. No adjustment set.

**Libraries**
```{r}
library(cmdstanr)
library(posterior)
library(tidyverse)
library(patchwork)
library(bayesplot)
library(ggplot2)
```
**Dataset**
must run chunk
```{r}
data <- read.csv("data.csv")
df <- data
#avoiding full 1
df <- df |>
  mutate(biofouling = if_else(
    biofouling == 0, 
    0,                                             
    (biofouling * (n() - 1) + 0.5) / n()           
  ))
# 0-1 zib can only handle 0-1
df$biofouling <- df$biofouling / 100

#2 most important zooplankton
df <- df %>%
  mutate(pred_zoo = copepoda + arrow_worm)
#temporary
df <- df %>% 
  mutate(nutrients =  NO2_um + NO3_um + PO4_um)

df <- df %>%
  rename(current = current_velocity_w17) %>%
  rename(seaweed_growth = blade_length) %>%
  rename(salinity = Salinity) %>%
  rename(precipitation = precipitation_mm_w17) %>%
  rename(daylight = daylight_duration_h_w17) %>%
  rename(sst = sea_surface_temperature_c_w17) %>%
  rename(phyto = total_phytoplankton) %>%
  rename(cyphonautes = cyphonaute) %>%
  rename(air_temp = avg_temp_w17)

#scaling
df <- df |>
  mutate(across(c(nutrients, phyto, seaweed_growth, current,
                  air_temp, sst, daylight, salinity,
                  pred_zoo, precipitation, cyphonautes),
                ~ scale(.)[,1]))

```

**Datalist**
```{r}
#Total list
obs_idx  <- which(!is.na(df$daylight))
miss_idx <- which(is.na(df$daylight))

daylight_total <- list(
  N = nrow(df),
  biofouling = df$biofouling,
  

  #exposure
  N_day_obs  = length(obs_idx),
  N_day_miss = length(miss_idx),
  day_obs_idx  = obs_idx,
  day_miss_idx = miss_idx,
  day_obs      = df$daylight[obs_idx],
  imp_day_mean = mean(df$daylight, na.rm = TRUE),
  imp_day_sd   = sd(df$daylight, na.rm = TRUE),

  #nested random effect - farm + replicate
  farm_id    = as.integer(factor(df$farm)),
  replicate_id = as.integer(factor(df$replicate)),
  J_farm     = length(unique(df$farm)),
  J_replicate  = length(unique(df$replicate)),

  # intervention grid
  N_interv = 11,
  do_daylight = seq(
    min(df$daylight, na.rm = TRUE),
    max(df$daylight, na.rm = TRUE),
    length.out = 11
  )
)
```

**Data list**
```{r}
#Direct (run chunk all at once)

#daylight mis/obs
day_obs_idx  <- which(!is.na(df$daylight))
day_miss_idx <- which(is.na(df$daylight))

#ccurrent miss/obs
cur_obs_idx  <- which(!is.na(df$current))
cur_miss_idx <- which(is.na(df$current))

#seaweed growth miss/obs
sw_obs_idx  <- which(!is.na(df$seaweed_growth))
sw_miss_idx <- which(is.na(df$seaweed_growth))

#phytoplanlton miss/obs
phy_obs_idx  <- which(!is.na(df$phyto))
phy_miss_idx <- which(is.na(df$phyto))

#list
daylight_direct <- list(
  N = nrow(df),
  biofouling = df$biofouling,

  #Exposure - daylight
  N_day_obs  = length(day_obs_idx),
  N_day_miss = length(day_miss_idx),
  day_obs_idx  = day_obs_idx,
  day_miss_idx = day_miss_idx,
  day_obs      = df$daylight[day_obs_idx],
  imp_day_mean = mean(df$daylight, na.rm = TRUE),
  imp_day_sd   = sd(df$daylight, na.rm = TRUE),

  #adjustment sets
  #current
  N_cur_obs  = length(cur_obs_idx),
  N_cur_miss = length(cur_miss_idx),
  cur_obs_idx  = cur_obs_idx,
  cur_miss_idx = cur_miss_idx,
  cur_obs      = df$current[cur_obs_idx],
  imp_cur_mean = mean(df$current, na.rm = TRUE),
  imp_cur_sd   = sd(df$current, na.rm = TRUE),

  #seaweed growth
  N_sw_obs  = length(sw_obs_idx),
  N_sw_miss = length(sw_miss_idx),
  sw_obs_idx  = sw_obs_idx,
  sw_miss_idx = sw_miss_idx,
  sw_obs      = df$seaweed_growth[sw_obs_idx],
  imp_sw_mean = mean(df$seaweed_growth, na.rm = TRUE),
  imp_sw_sd   = sd(df$seaweed_growth, na.rm = TRUE),

  #phytoplankton
  N_phy_obs  = length(phy_obs_idx),
  N_phy_miss = length(phy_miss_idx),
  phy_obs_idx  = phy_obs_idx,
  phy_miss_idx = phy_miss_idx,
  phy_obs      = df$phyto[phy_obs_idx],
  imp_phy_mean = mean(df$phyto, na.rm = TRUE),
  imp_phy_sd   = sd(df$phyto, na.rm = TRUE),

  #random effects
  #nested w/ farm -> replicates
  farm_id    = as.integer(factor(df$farm)),
  replicate_id = as.integer(factor(df$replicate)),
  J_farm     = length(unique(df$farm)),
  J_replicate  = length(unique(df$replicate)),

  #intervention grid
  N_interv = 11,
  do_daylight = seq(
    min(df$daylight, na.rm = TRUE),
    max(df$daylight, na.rm = TRUE),
    length.out = 11
  )
)
```


**Total**
```{r}
mod <- cmdstan_model("daylight.stan")
```

**Direct**
```{r}
moddirect <- cmdstan_model("daylightdirect.stan")
```


**Fitting - Total**
```{r}
#total
fit_total <- mod$sample(
  data = daylight_total,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.95,
  show_messages = TRUE
)
```
**Fitting - Direct**
```{r}
fit_direct <- moddirect$sample(
  data = daylight_direct,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.95,
  show_messages = TRUE
)
```


**Extracting**
```{r}
#sim
y_rep <- fit_total$draws("y_rep")
y_repd <- fit_direct$draws("y_rep")
y_repd_mat <- as_draws_matrix(y_repd)
y_rep_mat <- as_draws_matrix(y_rep) #convert to matrix
y_obs <- df$biofouling #observations

```

**Posterior Checks**

**ppc histogram (not great)**
```{r}
#total
pth <- ppc_hist(
  y = y_obs,
  yrep = y_rep_mat[1:50, ]  
) +
  ggplot2::labs(title = "Total Posterior predictive check - histogram")

#direct
pdh <- ppc_hist(
  y = y_obs,
  yrep = y_repd_mat[1:50, ]  
) +
  ggplot2::labs(title = "Direct Posterior predictive check - histogram")

pdh / pth #uglyy
```
**PPC no zeros sim vs true distribution**
```{r}
#no zero index
nz_idx <- y_obs > 0
#total plot
pt <- ppc_dens_overlay(
  y = y_obs[nz_idx],
  yrep = y_rep_mat[1:50, nz_idx]
) +
  ggplot2::labs(title = "Total PPC — non-zero values only")

#direct plot
pd <- ppc_dens_overlay(
  y = y_obs[nz_idx],
  yrep = y_repd_mat[1:50, nz_idx]
) +
  ggplot2::labs(title = "Direct PPC — non-zero values only")

pd / pt
```
**Zeroes observed**
```{r}
# Proportion of zeros in observed vs simulated
prop_zero <- function(x) mean(x == 0)

#pulls from y_rep_draws so will plot most recent model run 
ppc_stat(y_obs, y_rep_draws, stat = prop_zero) +
  labs(title = "Proportion of zeros: observed vs simulated")

```

**Farm distributions**
```{r}
df |>
  filter(biofouling > 0) |>
  ggplot(aes(x = biofouling, fill = farm)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~farm, scales = "free_y") +
  labs(title = "Non-zero biofouling distribution by farm")
```
**Mean and SD check**
```{r}
# SD + Mean
ppc_stat_2d(y_obs, y_rep_draws,
            stat = c("mean", "sd")) +
  labs(title = "Mean vs SD: observed vs simulated")
```
```{r}
ppc_stat(y_obs, y_rep_draws,
         stat = function(x) quantile(x, 0.95)) +
  labs(title = "95th percentile: observed vs simulated")
```

**Actual Results**

**model effects**
```{r}
b_int  <- fit_total$draws("b_daylight", format = "matrix") #intensity
b_zi   <- fit_total$draws("b_daylight_zi", format = "matrix") #occupancy

summary_effects <- function(x) {
  c(
    mean = mean(x),
    sd   = sd(x),
    q2.5 = quantile(x, 0.025),
    q97.5 = quantile(x, 0.975),
    prob_pos = mean(x > 0)
  )
}

summary_effects(b_int)
summary_effects(b_zi) #occupancy > intensity but both positive
```
**direct model effects**
```{r}
b_int  <- fit_direct$draws("b_daylight", format = "matrix") #intensity
b_zi   <- fit_direct$draws("b_daylight_zi", format = "matrix") #occupancy

summary_effects <- function(x) {
  c(
    mean = mean(x),
    sd   = sd(x),
    q2.5 = quantile(x, 0.025),
    q97.5 = quantile(x, 0.975),
    prob_pos = mean(x > 0)
  )
}

summary_effects(b_int)
summary_effects(b_zi) #Intensity higher than in total effect
```

**Gcomp**
```{r}
Ey_do <- fit_total$draws("Ey_do", format = "matrix")

Ey_summary <- data.frame(
  do_value = daylight_total$do_daylight,
  mean  = apply(Ey_do, 2, mean),
  lo    = apply(Ey_do, 2, quantile, 0.025),
  hi    = apply(Ey_do, 2, quantile, 0.975)
)


Ey_summary

#standard - increase light and biofouling increases
gtp <- ggplot(Ey_summary, aes(x = do_value, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  labs(
    title = "E[Y | do(daylight)] - G-computation",
    x = "Daylight intervention",
    y = "Expected biofouling"
  )
```
**Gcomp**
```{r}
Ey_do <- fit_direct$draws("Ey_do", format = "matrix")

Ey_summary <- data.frame(
  do_value = daylight_direct$do_daylight,
  mean  = apply(Ey_do, 2, mean),
  lo    = apply(Ey_do, 2, quantile, 0.025),
  hi    = apply(Ey_do, 2, quantile, 0.975)
)

Ey_summary

#standard - increase light and biofouling increases
gdp <- ggplot(Ey_summary, aes(x = do_value, y = mean)) +
  geom_line() +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  labs(
    title = "Direct E[Y | do(daylight)] - G-computation",
    x = "Daylight intervention",
    y = "Expected biofouling"
  )

gdp / gtp
```


