### Corresponding Rmd for SCM.stan
#### Started on 27/06/2026
#### Authored by 2654492b

=========================
This script is the wrapper for the Stan structural causal model (SCM) of bryozoan biofouling on cultivated kelp. It: 
  - Read and prepares the provided multi-farm dataset by selecting, renaming, and standardising the relevant predictors. 
  - Prepares each covariate for missing-data imputation by splitting into observed/missing indices.
  - Assembles stan data list
  - Fits model via cmdstanr then runs convergence diagnostics and posterior predictive checks.
  - Extracts structural path coefficients, total/direct causal contrasts, dose response curves, random effect SDs, and E-values.

*Required Packages**
```{r}
#cmdstanr - interface used to compile/run stan model (SCM.stan)
#posterior - for post convergence draws
#tidyverse - general use package
#patchwork - for combining multiple ggplot panels
#bayesplot - for ppc plotting 
#EValue - e value sensitivty analysis 

library(cmdstanr)
library(posterior)
library(tidyverse)
library(patchwork)
library(bayesplot)
library(ggplot2)
library(EValue)

```

**Dataset**
```{r}
#fully merged multi-farm dataset w/ all variables aggregated from appropriate windows
data <- read.csv("data.csv")
df <- data
```

**By class**

```{r}
#Option chunk, only run if interested in taxonomic phytoplankton class.
# inserts class level abundance rather than default total phytoplankton abundance
# requires win_14, the 0-14 day dataset for the phytoplankton classes


df <- df %>%
  left_join(
    win_14 %>%
      select(
        Identifier,
    Bacillariophyceae,
    Dinophyceae,
    Dictyochophyceae,
    Cryptophyceae,
    Prymnesiophyceae,
    Cyanophyceae
      ),
    by = "Identifier"
  )

```


**Data prep**
```{r}
#Zero inflated beta likelihoods cannot handle of 1
#Nudges non-zero values slightly off 1
df <- df |>
  mutate(biofouling = if_else(
    biofouling == 0, 
    0,                                             
    (biofouling * (n() - 1) + 0.5) / n()           
  ))

#Original values recorded as percentages (0-100%), need to rescale to 0-1 for beta likelihood
df$biofouling <- df$biofouling / 100

--- #Assigning predictors ---

#Combined predatory zooplankton from copepoda and arrow worms for a single exposure node
df <- df %>%
  mutate(pred_zoo = copepoda + arrow_worm)

#Combined dissolve nutrient concentrations
df <- df %>% 
  mutate(nutrients =  NO2_um + NO3_um + PO4_um)
#--Renaming names used by SCM--
df <- df %>%
  rename(current = current_velocity_w17) %>%
  rename(seaweed_growth = blade_length) %>%
  rename(salinity = Salinity) %>%
  rename(precipitation = precipitation_mm_w17) %>%
  rename(daylight = daylight_duration_h_w17) %>%
  rename(sst = sea_surface_temperature_c_w17) %>%
  rename(phyto = total_phytoplankton) %>% #change this to whatever phytoplankton class you want to check, otherwise leave as is
  rename(cyphonautes = cyphonaute) %>%
  rename(air_temp = avg_temp_w17) %>%
  rename(winter_nut = winter_nutrient_avg)


  
```


**Scaling Data**
```{r}
#standardising continous covariates
#path coefficients on a comparable scale for intervention grid
df <- df |>
  mutate(across(c(nutrients, phyto, seaweed_growth, current,
                  air_temp, sst, daylight, salinity,
                  pred_zoo, precipitation, cyphonautes, winter_nut ),
                ~ scale(.)[,1]))
# Integer indices for random effect group levels
df <- df |>
  mutate(
    farm_id      = as.integer(factor(farm)),
    replicate_id = as.integer(factor(replicate))
  )

#Farm weighting
#weight 1/n_farm then weights are rescaled to effective sample size
df <- df |> 
  group_by(farm) |>
  mutate(farm_weight = 1/ n()) |>
  ungroup()
df <- df |>
  mutate(farm_weight = farm_weight / sum(farm_weight) * n_distinct(farm))
```

**Prepping variables**
```{r}
#Splits each variable into its missing and observed values
prep_var <- function(x) {
  obs_idx  <- which(!is.na(x))
  miss_idx <- which(is.na(x))
  list(
    n_obs    = length(obs_idx),
    n_miss   = length(miss_idx),
    obs_idx  = obs_idx,
    miss_idx = miss_idx,
    obs_vals = x[obs_idx],
    imp_mean = mean(x, na.rm = TRUE),
    imp_sd   = max(sd(x, na.rm = TRUE), 1e-3)
  )
}

#posterior summary helper
summary_effects <- function(x) {
  c(
    mean = mean(x),
    sd = sd(x),
    q2.5 = quantile(x, 0.025),
    q97.5 = quantile(x, 0.975),
    prob_pos = mean(x > 0)
  )
}

#Pulls a vector-valued g-computation parameter out of fit and reshapes to do_value/mean/lo/hi dataframe for plotting
ey_to_df <- function(fit, prefix, grid_values) {
  vars <- paste0(prefix, "[", seq_along(grid_values), "]")
  fit$summary(vars) |>
    transmute(
      do_value = grid_values,
      mean = mean,
      lo = q5,
      hi = q95
    )
}

#Proportion of exact zeros in a vector
## PPC test statistic
prop_zero <- function(x) mean(x == 0)

```

**'Preparing' Variables**
```{r}
# Applying prep-var() to every covariate 
pv_airtemp <- prep_var(df$air_temp)
pv_current <- prep_var(df$current)
pv_daylight <- prep_var(df$daylight)
pv_precip <- prep_var(df$precipitation)
pv_predzoo <- prep_var(df$pred_zoo)
pv_sst <- prep_var(df$sst)
pv_sal <- prep_var(df$salinity)
pv_nut <- prep_var(df$nutrients)
pv_seaweed <- prep_var(df$seaweed_growth)
pv_phyto <- prep_var(df$phyto)
pv_cyph <- prep_var(df$cyphonautes)
pv_winter_nut <- prep_var(df$winter_nut)
```

**Intervention grid**
```{r}
#grid of intervention values used for every do(X = x) query
N_interv <- 11
interv_grid <- seq(-1, 1, length.out = N_interv) #range of interventions
```



**Model datalist**
```{r}

#Full datalist to be passed to Stan
stan_data <- list(
  N = nrow(df),
  #random effects
  J_farm = max(df$farm_id),
  J_replicate = max(df$replicate_id),
  farm_id = df$farm_id,
  replicate_id = df$replicate_id,
  farm_weight = df$farm_weight, 
  
  #response
  #zero-inflated beta response
  biofouling = df$biofouling,
  
  #exposure nodes -
  #starting w/ root nodes
  
  #Air Temperature (intermediate: daylight -> air_temp)
  N_airtemp_obs = pv_airtemp$n_obs,
  N_airtemp_miss = pv_airtemp$n_miss,
  airtemp_obs_idx = pv_airtemp$obs_idx,
  airtemp_miss_idx = pv_airtemp$miss_idx,
  airtemp_obs = pv_airtemp$obs_vals,
  
  #winter nutrients
  N_winternut_obs = pv_winter_nut$n_obs,
  N_winternut_miss = pv_winter_nut$n_miss,
  winternut_obs_idx = pv_winter_nut$obs_idx,
  winternut_miss_idx = pv_winter_nut$miss_idx,
  winternut_obs = pv_winter_nut$obs_vals,
  imp_winternut_mean = pv_winter_nut$imp_mean,
  imp_winternut_sd = pv_winter_nut$imp_sd,
  
  #winter nutrients
  
  
  #Daylight (root node)
  N_daylight_obs = pv_daylight$n_obs,
  N_daylight_miss = pv_daylight$n_miss,
  daylight_obs_idx = pv_daylight$obs_idx,
  daylight_miss_idx = pv_daylight$miss_idx,
  daylight_obs = pv_daylight$obs_vals,
  imp_daylight_mean = pv_daylight$imp_mean,
  imp_daylight_sd = pv_daylight$imp_sd,
  
  #Current
  N_current_obs = pv_current$n_obs,
  N_current_miss = pv_current$n_miss,
  current_obs_idx = pv_current$obs_idx,
  current_miss_idx = pv_current$miss_idx, 
  current_obs = pv_current$obs_vals,
  imp_current_mean = pv_current$imp_mean, 
  imp_current_sd = pv_current$imp_sd,
  
  #Precipitation 
  N_precip_obs = pv_precip$n_obs,
  N_precip_miss = pv_precip$n_miss,
  precip_obs_idx = pv_precip$obs_idx,
  precip_miss_idx = pv_precip$miss_idx,
  precip_obs = pv_precip$obs_vals, 
  imp_precip_mean = pv_precip$imp_mean,
  imp_precip_sd = pv_precip$imp_sd,
  
  #Predatory zooplankton
  N_predzoo_obs = pv_predzoo$n_obs,
  N_predzoo_miss = pv_predzoo$n_miss,
  predzoo_obs_idx = pv_predzoo$obs_idx,
  predzoo_miss_idx = pv_predzoo$miss_idx,
  imp_predzoo_mean = pv_predzoo$imp_mean,
  imp_predzoo_sd = pv_predzoo$imp_sd,
  predzoo_obs = pv_predzoo$obs_vals,
  
  #Sea surface temperature 
  N_sst_obs = pv_sst$n_obs,
  N_sst_miss = pv_sst$n_miss,
  sst_obs_idx = pv_sst$obs_idx,
  sst_miss_idx = pv_sst$miss_idx,
  sst_obs = pv_sst$obs_vals,
  
  #Salinity
  N_sal_obs = pv_sal$n_obs,
  N_sal_miss = pv_sal$n_miss,
  sal_obs_idx = pv_sal$obs_idx,
  sal_miss_idx = pv_sal$miss_idx,
  sal_obs = pv_sal$obs_vals,
  
  #Nutrients - NO3 + NO2 + PO4
  N_nut_obs = pv_nut$n_obs,
  N_nut_miss = pv_nut$n_miss,
  nut_obs_idx = pv_nut$obs_idx,
  nut_miss_idx = pv_nut$miss_idx,
  nut_obs = pv_nut$obs_vals,
  
  #Seaweed growth 
  N_seaweed_obs = pv_seaweed$n_obs,
  N_seaweed_miss = pv_seaweed$n_miss,
  seaweed_obs_idx = pv_seaweed$obs_idx,
  seaweed_miss_idx = pv_seaweed$miss_idx,
  seaweed_obs = pv_seaweed$obs_vals,
  
  #Phytoplankton
  N_phyto_obs = pv_phyto$n_obs,
  N_phyto_miss = pv_phyto$n_miss,
  phyto_obs_idx = pv_phyto$obs_idx,
  phyto_miss_idx = pv_phyto$miss_idx,
  phyto_obs = pv_phyto$obs_vals,
  
  #Cyphonautes
  N_cyph_obs = pv_cyph$n_obs,
  N_cyph_miss = pv_cyph$n_miss,
  cyph_obs_idx = pv_cyph$obs_idx,
  cyph_miss_idx = pv_cyph$miss_idx,
  cyph_obs = pv_cyph$obs_vals,
  
  #Intervention grids - carrying over gcomp stuff from causal model
  N_interv = N_interv,
  do_airtemp = interv_grid,
  do_current = interv_grid,
  do_daylight = interv_grid,
  do_precip = interv_grid,
  do_predzoo = interv_grid,
  do_sst = interv_grid,
  do_nutrients = interv_grid,
  do_seaweed = interv_grid,
  do_phyto = interv_grid,
  do_salinity = interv_grid,
  do_cypho = interv_grid,
  do_winternut = interv_grid
) 
```


**Compiling model**
```{r}
#Compile stan program
model <- cmdstan_model("SCMre.stan")
```
**Fitting model**
```{r}
fit <- model$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 2000,
  iter_sampling = 5000,
  adapt_delta = 0.95,
  max_treedepth = 12,
  refresh = 100,
  show_messages = TRUE
)

```


**Convergence diagnostics**
```{r}
#MCMC health checks
# divergences
#max rhat (< 1.01)
# min ESS bulk (> 400)
fit_summary <- fit$summary()
fit_diag <- fit$sampler_diagnostics(format = "draws_df")

tibble(
  divergences = sum(fit_diag$divergent__ == 1, na.rm = TRUE),
  max_rhat = max(fit_summary$rhat, na.rm = TRUE),
  min_ess_bulk = min(fit_summary$ess_bulk, na.rm = TRUE),
  n_bad_rhat = sum(fit_summary$rhat > 1.01, na.rm = TRUE),
  n_low_ess = sum(fit_summary$ess_bulk < 400, na.rm = TRUE)
)

#If variables are problematic, which are they?
fit_summary |>
  filter(rhat > 1.01 | ess_bulk < 400) |>
  arrange(desc(rhat)) |>
  select(variable, mean, sd, rhat, ess_bulk) |>
  head(20)
```


**Posterior predictive checks — SCM**
```{r}
#drawing y_rep and comparing against observed y to check model fit
y_obs <- df$biofouling
y_rep_mat <- fit$draws("y_rep", format = "matrix")
y_rep_draws <- posterior::as_draws_matrix(fit$draws("y_rep"))
n_ppc_draws <- min(50, nrow(y_rep_mat))


# overall distribution check
p_ppc_hist <- ppc_hist(
  y = y_obs,
  yrep = y_rep_mat[seq_len(n_ppc_draws), ]
) +
  labs(title = "SCM — posterior predictive histogram")

#non-zero observations to check beta component shape
nz_idx <- y_obs > 0
p_ppc_dens <- ppc_dens_overlay(
  y = y_obs[nz_idx],
  yrep = y_rep_mat[seq_len(n_ppc_draws), nz_idx]
) +
  labs(title = "SCM — non-zero values only")

#does the model reproduce the observed proportion of true zero?
p_ppc_zero <- ppc_stat(y_obs, y_rep_draws, stat = prop_zero) +
  labs(title = "SCM — proportion of zeros")

#mean and SD reproduction
p_ppc_mean_sd <- ppc_stat_2d(y_obs, y_rep_draws, stat = c("mean", "sd")) +
  labs(title = "SCM — mean vs SD")

p_ppc_hist / p_ppc_dens
p_ppc_zero
p_ppc_mean_sd
```

**Structural coefficients**
```{r}
#Main structural path coefficients
fit$summary(c(
  "b_airtemp_daylight", "b_sst_airtemp", "b_seaweed_daylight",
  "b_phyto_daylight", "b_bf_phyto", "b_bf_seaweed", "b_bf_cyph",
  "b_zi_seaweed", "sigma_farm", "phi", "b_phyto_winternut", "b_seaweed_winternut"
)) |>
  select(variable, mean, sd, rhat, ess_bulk)
```
**extracting gcomp summary**
```{r}
#Labels for each exposure, used in future plots
exposure_map <- c(
  airtemp   = "Air temperature",
  current   = "Current",
  daylight  = "Daylight",
  precip    = "Precipitation",
  predzoo   = "Pred. zooplankton",
  sst       = "SST",
  nutrients = "Nutrients",
  seaweed   = "Seaweed growth",
  phyto     = "Phytoplankton",
  salinity  = "Salinity",
  cypho     = "Cyphonautes",
  winternut = "Winter nutrients"
)
#Extracting generated quantity Ey_do<exposure> 
#forming mean + 95% CI at each grid point
#total causal effect summary
gcomp_summary <- imap_dfr(exposure_map, function(label, exp) {
  param <- paste0("Ey_do_", exp)
  draws <- fit$draws(param, format = "matrix")
  map_dfr(seq_len(N_interv), function(k) {
    col <- draws[, k]
    tibble(
      exposure = exp,
      label    = label,
      do_value = interv_grid[k],
      Ey_mean  = mean(col),
      Ey_lo    = quantile(col, 0.025),
      Ey_hi    = quantile(col, 0.975)
    )
  })
})
```


**Extracting contrast summaries**
```{r}
exposures <- c("airtemp", "current", "daylight", "precip",
               "predzoo", "sst", "nutrients", "seaweed",
               "phyto", "salinity", "cypho", "winternut")

#Total causal effect of each exposure
#contrast between top and bottom ends of intervention grid
contrast_summary <- map_dfr(exposures, function(exp) {
  param    <- paste0("Ey_do_", exp)
  draws    <- fit$draws(param, format = "matrix")
  contrast <- draws[, N_interv] - draws[, 1]
  tibble(
    exposure      = exp,
    contrast_mean = mean(contrast),
    contrast_lo   = quantile(contrast, 0.025),
    contrast_hi   = quantile(contrast, 0.975),
    prob_pos      = mean(contrast > 0)
  )
}) |>
  distinct(exposure, .keep_all = TRUE)
```

**Direct Contrasts**
```{r}
#subset of exposures for which stan model also computes a _direct generated quantity for
exposures_direct <- c("daylight", "current", "sst", "phyto")

contrast_direct <- map_dfr(exposures_direct, function(exp) {
  param    <- paste0("Ey_do_", exp, "_direct")
  draws    <- fit$draws(param, format = "matrix")
  contrast <- draws[, N_interv] - draws[, 1]
  tibble(
    exposure      = exp,
    estimand      = "direct",
    contrast_mean = mean(contrast),
    contrast_lo   = quantile(contrast, 0.025),
    contrast_hi   = quantile(contrast, 0.975),
    prob_pos      = mean(contrast > 0)
  )
})
```

**Contrast Summary Graph**
```{r}


#Forest style plot of total causal contrasts for every exposure
#Coloured by significant, grey = non significant, orange = significant. 
contrast_summary |>
  mutate(
    sig = (contrast_lo > 0 | contrast_hi < 0),
    exposure = recode(exposure,
      airtemp   = "Air Temperature",
      current   = "Current",
      daylight  = "Daylight Hours",
      precip    = "Precipitation",
      predzoo   = "Pred. Zooplankton",
      sst       = "SST",
      nutrients = "Nutrients",
      seaweed   = "Seaweed Growth",
      phyto     = "Phytoplankton",
      salinity  = "Salinity",
      cypho     = "Cyphonautes",
      winter_nut = "Winter Nutrients"
    )
  ) |>
  ggplot(aes(x = contrast_mean * 100,
             xmin = contrast_lo * 100,
             xmax = contrast_hi * 100,
             y = reorder(exposure, contrast_mean),
             colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(height = 0.3) +
  geom_point(size = 3) +
  scale_colour_manual(values = c("grey60", "#D55E00"),
                      labels = c("CI includes 0", "CI excludes 0")) +
  labs(title  = "Total causal contrasts: do(X = +1SD) - do(X = −1SD)",
       x      = "Change in expected % bryozoan cover (± 95% CI)",
       y      = NULL,
       colour = NULL) +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")

contrast_summary <- contrast_summary |>
  distinct(exposure, .keep_all = TRUE)

unique(contrast_summary$exposure)

```

**Total and direct effect summaries**
```{r}

#Combining total and direct effects for comparasion
contrast_both <- bind_rows(
  contrast_summary |>
    filter(exposure %in% exposures_direct) |>
    mutate(estimand = "total"),
  contrast_direct
) |>
  mutate(
    sig = (contrast_lo > 0 | contrast_hi < 0),
    exposure = recode(exposure,
      daylight = "Daylight Hours",
      current  = "Current",
      sst      = "SST",
      phyto    = "Phytoplankton"
    )
  )

ggplot(contrast_both,
       aes(x = contrast_mean * 100,
           xmin = contrast_lo * 100,
           xmax = contrast_hi * 100,
           y = exposure,
           colour = estimand,
           alpha = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(height = 0.25,
                 position = position_dodge(width = 0.5)) +
  geom_point(size = 3,
             position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c(direct = "#0072B2", total = "#E69F00")) +
  scale_alpha_manual(values = c("FALSE" = 0.3, "TRUE" = 1),
                     guide  = "none") +
  labs(title   = "Total vs direct causal contrasts",
       subtitle = "Difference = indirect effect through mediators",
       x        = "Change in expected % cover (± 95% CI)",
       y        = NULL,
       colour   = "Estimand") +
  theme_bw(base_size = 12) +
  theme(legend.position = "bottom")



```

**Intervention curves — SCM (all exposures)**
```{r}
exposure_map <- c(
  Ey_do_daylight = "Daylight",
  Ey_do_airtemp = "Air temperature",
  Ey_do_sst = "SST",
  Ey_do_current = "Current",
  Ey_do_precip = "Precipitation",
  Ey_do_nutrients = "Nutrients",
  Ey_do_seaweed = "Seaweed growth",
  Ey_do_phyto = "Phytoplankton",
  Ey_do_salinity = "Salinity",
  Ey_do_predzoo = "Predatory zooplankton",
  Ey_do_cypho = "Cyphonautes",
  Ey_do_daylight_direct = "Direct Daylight",
  Ey_do_current_direct = "Direct Current",
  Ey_do_sst_direct = "Direct SST",
  Ey_do_phyto_direct = "Phytoplankton Direct",
  Ey_do_winternut = "Winter Nutrients"
)

#full dose-response table
Ey_scm_all <- imap_dfr(exposure_map, \(label, prefix) {
  ey_to_df(fit, prefix, interv_grid) |>
    mutate(exposure = label)
})

#faceted dose-response curves
#expected biofouling as function of intervention level
#one panel per exposure
p_interv_all <- ggplot(Ey_scm_all, aes(x = do_value, y = mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~exposure, scales = "free_y", ncol = 3) +
  labs(
    title = expression(SCM ~ E[Y * "|" * do(X)]),
    x = "Intervention level (standardised)",
    y = "Expected biofouling"
  ) +
  theme_bw()

p_interv_all
```

**Pathways**
```{r}
# All path coefficients in the DAG
path_params <- c(
  # SST sub-model
  "b_sst_airtemp",
  # Salinity sub-model
  "b_sal_precip",
  # Nutrients sub-model
  "b_nut_sst", "b_nut_current", "b_nut_precip", "b_nut_phyto",
  # Seaweed sub-model
  "b_seaweed_nut", "b_seaweed_daylight", "b_seaweed_winternut",
  # Phyto sub-model
  "b_phyto_sal", "b_phyto_winternut", "b_phyto_daylight",
  # Cyphonautes sub-model
  "b_cyph_phyto", "b_cyph_predzoo", "b_cyph_current",
  # Biofouling intensity
  "b_bf_seaweed", "b_bf_phyto", "b_bf_cyph",
  # Biofouling occupancy
  "b_zi_seaweed", "b_zi_phyto", "b_zi_cyph",
  #air
  "b_airtemp_daylight"
)

#posterior mean/sd/95% ci/ prob_pos for every path coefficient
path_summary <- map_dfr(path_params, function(p) {
  draws <- fit$draws(p, format = "matrix")
  tibble(
    parameter = p,
    mean      = mean(draws),
    sd        = sd(draws),
    q2.5      = quantile(draws, 0.025),
    q97.5     = quantile(draws, 0.975),
    prob_pos  = mean(draws > 0)
  )
})



```

**Pathway summary graph**
```{r}
#classifies every path coefficient into its submodel for facetting 
#relables as x -> y.
#plots every coefficient as a forest plot faceted by submodel
p_paths <- path_summary |>
  mutate(
    sig      = (q2.5 > 0 | q97.5 < 0),
    submodel = case_when(
      str_starts(parameter, "b_airtemp")  ~ "Air Temperature",
      str_starts(parameter, "b_sst")      ~ "SST",
      str_starts(parameter, "b_sal")      ~ "Salinity",
      str_starts(parameter, "b_nut")      ~ "Nutrients",
      str_starts(parameter, "b_seaweed")  ~ "Seaweed Growth",
      str_starts(parameter, "b_phyto")    ~ "Phytoplankton",
      str_starts(parameter, "b_cyph")     ~ "Cyphonautes",
      str_starts(parameter, "b_bf")       ~ "Biofouling (Intensity)",
      str_starts(parameter, "b_zi")       ~ "Biofouling (Occupancy)"
    ),
    submodel = factor(submodel, levels = c(
      "Biofouling (Intensity)",
      "Biofouling (Occupancy)",
      "Cyphonautes",
      "Nutrients",
      "Phytoplankton",
      "Salinity",
      "Seaweed Growth",
      "Air Temperature",
      "SST"
    )),
    parameter = recode(parameter,
      b_airtemp_daylight  = "Daylight → Air temp",
      b_sst_airtemp       = "Air temp → SST",
      b_sal_precip        = "Precipitation → Salinity",
      b_nut_sst           = "SST → Nutrients",
      b_nut_current       = "Current → Nutrients",
      b_nut_precip        = "Precipitation → Nutrients",
      b_seaweed_nut       = "Nutrients → Seaweed",
      b_seaweed_daylight  = "Daylight → Seaweed",
      b_phyto_sal         = "Salinity → Phyto",
      b_phyto_nut         = "Nutrients → Phyto",
      b_phyto_daylight    = "Daylight → Phyto",
      b_cyph_phyto        = "Phytoplankton → Cyphonautes",
      b_cyph_predzoo      = "Pred. zoo → Cyphonautes",
      b_cyph_current      = "Current → Cyphonautes",
      b_bf_seaweed        = "Seaweed → Biofouling",
      b_bf_phyto          = "Phytoplankton → Biofouling",
      b_bf_cyph           = "Cyphonautes → Biofouling",
      b_zi_seaweed        = "Seaweed → Occupancy",
      b_zi_phyto          = "Phytoplankton → Occupancy",
      b_zi_cyph           = "Cyphonautes → Occupancy",
      b_phyto_winternut   = "Winter Nutrients → Phytoplankton",
      b_nut_phyto         = "Phytoplankton → Nutrients",
      b_seaweed_winternut = "Winter Nutrients → Seaweed Growth"
    )
  ) |>
  ggplot(aes(x = mean, xmin = q2.5, xmax = q97.5,
             y = reorder(parameter, mean),
             colour = sig)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_errorbar(height = 0.3) +
  geom_point(size = 2.5) +
  facet_wrap(~submodel, scales = "free_y", ncol = 2) +
  scale_colour_manual(values = c("grey60", "#D55E00"),
                      labels = c("CI includes 0", "CI excludes 0")) +
  labs(title   = "Structural path coefficients",
       x       = "Posterior mean (± 95% CI)",
       y       = NULL,
       colour  = NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom")

p_paths


```

**Random effect graphs**
```{r}
#posterior SDs of two random effect groups
re_draws <- map_dfr(c("sigma_farm", "sigma_replicate"), function(p) {
  draws <- fit$draws(p, format = "matrix")
  tibble(
    parameter = p,
    mean      = mean(draws),
    q2.5      = quantile(draws, 0.025),
    q97.5     = quantile(draws, 0.975)
  )
}) |>
  mutate(parameter = recode(parameter,
    sigma_farm      = "Between-farm SD",
    sigma_replicate = "Between-blade SD"
  ))

farm_re<- ggplot(re_draws,
       aes(x = mean, xmin = q2.5, xmax = q97.5,
           y = parameter)) +
  geom_errorbarh(height = 0.2) +
  geom_point(size = 3, colour = "steelblue") +
  labs(title = "Random effect standard deviations",
       x     = "Posterior mean (± 95% CI)",
       y     = NULL) +
  theme_bw(base_size = 12)


```






**E-values**
```{r}
#Sensitivity analysis
#Evalues
ev_rd <- gcomp_summary |>
  group_by(exposure) |>
  summarise(
    Ey_lo = Ey_mean[which.min(do_value)],
    Ey_hi = Ey_mean[which.max(do_value)],
    .groups = "drop"
  ) |>
  mutate(
    RD        = Ey_hi - Ey_lo,
    RD_abs    = abs(RD),
    Ey_mid    = (Ey_lo + Ey_hi) / 2,
    RR_mid    = (Ey_mid + RD_abs/2) / pmax(Ey_mid - RD_abs/2, 1e-4),
    RR_mid    = ifelse(RR_mid < 1, 1/RR_mid, RR_mid),
    e_value   = ifelse(
      RR_mid <= 1, 1,
      RR_mid + sqrt(RR_mid * (RR_mid - 1))
    ),
    unstable  = Ey_lo < 0.005
  ) |>
  select(exposure, RD, Ey_mid, RR_mid, e_value, unstable)

print(ev_rd)


```

