### Corresponding Rmd for SCM.stan
#### Started on 27/06/2026

Data formatting, model fitting, convergence checks, posterior predictive checks,
coefficient summaries, intervention curves, and comparison with the daylight-only benchmark.
```{r}
library(cmdstanr)
library(posterior)
library(tidyverse)
library(patchwork)
library(bayesplot)
library(ggplot2)
library(EValue)
```

**Dataset**
```{r}
data <- read.csv("data.csv")
df <- data
#avoiding full 1
df <- df |>
  mutate(biofouling = if_else(
    biofouling == 0, 
    0,                                             
    (biofouling * (n() - 1) + 0.5) / n()           
  ))
# 0-1 zib can only handle 0-1
df$biofouling <- df$biofouling / 100

#2 most important zooplankton
df <- df %>%
  mutate(pred_zoo = copepoda + arrow_worm)
#temporary
df <- df %>% 
  mutate(nutrients =  NO2_um + NO3_um + PO4_um)

df <- df %>%
  rename(current = current_velocity_w17) %>%
  rename(seaweed_growth = blade_length) %>%
  rename(salinity = Salinity) %>%
  rename(precipitation = precipitation_mm_w17) %>%
  rename(daylight = daylight_duration_h_w17) %>%
  rename(sst = sea_surface_temperature_c_w17) %>%
  rename(phyto = total_phytoplankton) %>%
  rename(cyphonautes = cyphonaute) %>%
  rename(air_temp = avg_temp_w17)

#scaling
df <- df |>
  mutate(across(c(nutrients, phyto, seaweed_growth, current,
                  air_temp, sst, daylight, salinity,
                  pred_zoo, precipitation, cyphonautes),
                ~ scale(.)[,1]))

df <- df |>
  mutate(
    farm_id      = as.integer(factor(farm)),
    replicate_id = as.integer(factor(replicate))
  )

```

```{r}
#Splits each variable into its missing and observed values
#helps with imputation
prep_var <- function(x) {
  obs_idx  <- which(!is.na(x))
  miss_idx <- which(is.na(x))
  list(
    n_obs    = length(obs_idx),
    n_miss   = length(miss_idx),
    obs_idx  = obs_idx,
    miss_idx = miss_idx,
    obs_vals = x[obs_idx],
    imp_mean = mean(x, na.rm = TRUE),
    imp_sd   = max(sd(x, na.rm = TRUE), 1e-3)
  )
}

summary_effects <- function(x) {
  c(
    mean = mean(x),
    sd = sd(x),
    q2.5 = quantile(x, 0.025),
    q97.5 = quantile(x, 0.975),
    prob_pos = mean(x > 0)
  )
}

ey_to_df <- function(fit, prefix, grid_values) {
  vars <- paste0(prefix, "[", seq_along(grid_values), "]")
  fit$summary(vars) |>
    transmute(
      do_value = grid_values,
      mean = mean,
      lo = q5,
      hi = q95
    )
}

prop_zero <- function(x) mean(x == 0)

```

**'Preparing' Variables**
```{r}
pv_airtemp <- prep_var(df$air_temp)
pv_current <- prep_var(df$current)
pv_daylight <- prep_var(df$daylight)
pv_precip <- prep_var(df$precipitation)
pv_predzoo <- prep_var(df$pred_zoo)
pv_sst <- prep_var(df$sst)
pv_sal <- prep_var(df$salinity)
pv_nut <- prep_var(df$nutrients)
pv_seaweed <- prep_var(df$seaweed_growth)
pv_phyto <- prep_var(df$phyto)
pv_cyph <- prep_var(df$cyphonautes)
```

**Intervention grid shtuff**
```{r}
N_interv <- 11 
interv_grid <- seq(-1, 1, length.out = N_interv) #range of interventions
```



**Model datalist**
```{r}
stan_data <- list(
  N = nrow(df),
  #random effectss
  J_farm = max(df$farm_id),
  J_replicate = max(df$replicate_id),
  farm_id = df$farm_id,
  replicate_id = df$replicate_id,
  
  #response
  biofouling = df$biofouling,
  
  #exposure nodes -
  #starting w/ root nodes
  
  #Air Temperature (intermediate: daylight -> air_temp)
  N_airtemp_obs = pv_airtemp$n_obs,
  N_airtemp_miss = pv_airtemp$n_miss,
  airtemp_obs_idx = pv_airtemp$obs_idx,
  airtemp_miss_idx = pv_airtemp$miss_idx,
  airtemp_obs = pv_airtemp$obs_vals,
  
  #Daylight (root node)
  N_daylight_obs = pv_daylight$n_obs,
  N_daylight_miss = pv_daylight$n_miss,
  daylight_obs_idx = pv_daylight$obs_idx,
  daylight_miss_idx = pv_daylight$miss_idx,
  daylight_obs = pv_daylight$obs_vals,
  imp_daylight_mean = pv_daylight$imp_mean,
  imp_daylight_sd = pv_daylight$imp_sd,
  
  #Current
  N_current_obs = pv_current$n_obs,
  N_current_miss = pv_current$n_miss,
  current_obs_idx = pv_current$obs_idx,
  current_miss_idx = pv_current$miss_idx, 
  current_obs = pv_current$obs_vals,
  imp_current_mean = pv_current$imp_mean, 
  imp_current_sd = pv_current$imp_sd,
  
  #Precipitation 
  N_precip_obs = pv_precip$n_obs,
  N_precip_miss = pv_precip$n_miss,
  precip_obs_idx = pv_precip$obs_idx,
  precip_miss_idx = pv_precip$miss_idx,
  precip_obs = pv_precip$obs_vals, 
  imp_precip_mean = pv_precip$imp_mean,
  imp_precip_sd = pv_precip$imp_sd,
  
  #Predatory zooplankton
  N_predzoo_obs = pv_predzoo$n_obs,
  N_predzoo_miss = pv_predzoo$n_miss,
  predzoo_obs_idx = pv_predzoo$obs_idx,
  predzoo_miss_idx = pv_predzoo$miss_idx,
  imp_predzoo_mean = pv_predzoo$imp_mean,
  imp_predzoo_sd = pv_predzoo$imp_sd,
  predzoo_obs = pv_predzoo$obs_vals,
  
  #Sea surface temperature 
  N_sst_obs = pv_sst$n_obs,
  N_sst_miss = pv_sst$n_miss,
  sst_obs_idx = pv_sst$obs_idx,
  sst_miss_idx = pv_sst$miss_idx,
  sst_obs = pv_sst$obs_vals,
  
  #Salinity
  N_sal_obs = pv_sal$n_obs,
  N_sal_miss = pv_sal$n_miss,
  sal_obs_idx = pv_sal$obs_idx,
  sal_miss_idx = pv_sal$miss_idx,
  sal_obs = pv_sal$obs_vals,
  
  #Nutrients - NO3 + NO2 + PO4
  N_nut_obs = pv_nut$n_obs,
  N_nut_miss = pv_nut$n_miss,
  nut_obs_idx = pv_nut$obs_idx,
  nut_miss_idx = pv_nut$miss_idx,
  nut_obs = pv_nut$obs_vals,
  
  #Seaweed growth 
  N_seaweed_obs = pv_seaweed$n_obs,
  N_seaweed_miss = pv_seaweed$n_miss,
  seaweed_obs_idx = pv_seaweed$obs_idx,
  seaweed_miss_idx = pv_seaweed$miss_idx,
  seaweed_obs = pv_seaweed$obs_vals,
  
  #Phytoplankton
  N_phyto_obs = pv_phyto$n_obs,
  N_phyto_miss = pv_phyto$n_miss,
  phyto_obs_idx = pv_phyto$obs_idx,
  phyto_miss_idx = pv_phyto$miss_idx,
  phyto_obs = pv_phyto$obs_vals,
  
  #Cyphonautes
  N_cyph_obs = pv_cyph$n_obs,
  N_cyph_miss = pv_cyph$n_miss,
  cyph_obs_idx = pv_cyph$obs_idx,
  cyph_miss_idx = pv_cyph$miss_idx,
  cyph_obs = pv_cyph$obs_vals,
  
  #Intervention grids - carrying over gcomp stuff from causal model
  N_interv = N_interv,
  do_airtemp = interv_grid,
  do_current = interv_grid,
  do_daylight = interv_grid,
  do_precip = interv_grid,
  do_predzoo = interv_grid,
  do_sst = interv_grid,
  do_nutrients = interv_grid,
  do_seaweed = interv_grid,
  do_phyto = interv_grid,
  do_salinity = interv_grid,
  do_cypho = interv_grid
) 
```


**Compiling model**
```{r}
model <- cmdstan_model("SCM.stan")
```
**Fitting model**
```{r}
fit <- model$sample(
  data = stan_data,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.95,
  max_treedepth = 12,
  refresh = 100,
  show_messages = TRUE
)

```


**Convergence diagnostics**
```{r}
fit_summary <- fit$summary()
fit_diag <- fit$sampler_diagnostics(format = "draws_df")

tibble(
  divergences = sum(fit_diag$divergent__ == 1, na.rm = TRUE),
  max_rhat = max(fit_summary$rhat, na.rm = TRUE),
  min_ess_bulk = min(fit_summary$ess_bulk, na.rm = TRUE),
  n_bad_rhat = sum(fit_summary$rhat > 1.01, na.rm = TRUE),
  n_low_ess = sum(fit_summary$ess_bulk < 400, na.rm = TRUE)
)

fit_summary |>
  filter(rhat > 1.01 | ess_bulk < 400) |>
  arrange(desc(rhat)) |>
  select(variable, mean, sd, rhat, ess_bulk) |>
  head(20)
```


**Posterior predictive checks — SCM**
```{r}
y_obs <- df$biofouling
y_rep_mat <- fit$draws("y_rep", format = "matrix")
y_rep_draws <- posterior::as_draws_matrix(fit$draws("y_rep"))
n_ppc_draws <- min(50, nrow(y_rep_mat))

p_ppc_hist <- ppc_hist(
  y = y_obs,
  yrep = y_rep_mat[seq_len(n_ppc_draws), ]
) +
  labs(title = "SCM — posterior predictive histogram")

nz_idx <- y_obs > 0
p_ppc_dens <- ppc_dens_overlay(
  y = y_obs[nz_idx],
  yrep = y_rep_mat[seq_len(n_ppc_draws), nz_idx]
) +
  labs(title = "SCM — non-zero values only")

p_ppc_zero <- ppc_stat(y_obs, y_rep_draws, stat = prop_zero) +
  labs(title = "SCM — proportion of zeros")

p_ppc_mean_sd <- ppc_stat_2d(y_obs, y_rep_draws, stat = c("mean", "sd")) +
  labs(title = "SCM — mean vs SD")

p_ppc_hist / p_ppc_dens
p_ppc_zero
p_ppc_mean_sd
```


**Structural coefficients**
```{r}
fit$summary(c(
  "b_airtemp_daylight", "b_sst_airtemp", "b_seaweed_daylight",
  "b_phyto_daylight", "b_bf_phyto", "b_bf_seaweed", "b_bf_cyph",
  "b_zi_seaweed", "sigma_farm", "phi"
)) |>
  select(variable, mean, sd, rhat, ess_bulk)
```


**Intervention curves — SCM (all exposures)**
```{r}
exposure_map <- c(
  Ey_do_daylight = "Daylight",
  Ey_do_airtemp = "Air temperature",
  Ey_do_sst = "SST",
  Ey_do_current = "Current",
  Ey_do_precip = "Precipitation",
  Ey_do_nutrients = "Nutrients",
  Ey_do_seaweed = "Seaweed growth",
  Ey_do_phyto = "Phytoplankton",
  Ey_do_salinity = "Salinity",
  Ey_do_predzoo = "Predatory zooplankton",
  Ey_do_cypho = "Cyphonautes"
)

Ey_scm_all <- imap_dfr(exposure_map, \(label, prefix) {
  ey_to_df(fit, prefix, interv_grid) |>
    mutate(exposure = label)
})

p_interv_all <- ggplot(Ey_scm_all, aes(x = do_value, y = mean)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~exposure, scales = "free_y", ncol = 3) +
  labs(
    title = expression(SCM ~ E[Y * "|" * do(X)]),
    x = "Intervention level (standardised)",
    y = "Expected biofouling"
  ) +
  theme_bw()

p_interv_all
```


**Daylight-only benchmark**
```{r}
daylight_total <- list(
  N = nrow(df),
  biofouling = df$biofouling,
  N_day_obs = pv_daylight$n_obs,
  N_day_miss = pv_daylight$n_miss,
  day_obs_idx = pv_daylight$obs_idx,
  day_miss_idx = pv_daylight$miss_idx,
  day_obs = pv_daylight$obs_vals,
  imp_day_mean = pv_daylight$imp_mean,
  imp_day_sd = pv_daylight$imp_sd,
  farm_id = df$farm_id,
  replicate_id = df$replicate_id,
  J_farm = max(df$farm_id),
  J_replicate = max(df$replicate_id),
  N_interv = N_interv,
  do_daylight = interv_grid
)

mod_daylight <- cmdstan_model("daylight.stan")
fit_daylight <- mod_daylight$sample(
  data = daylight_total,
  chains = 4,
  parallel_chains = 4,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.95,
  max_treedepth = 12,
  refresh = 100,
  show_messages = TRUE
)
```


**Posterior predictive checks — daylight-only benchmark**
```{r}
y_rep_day_mat <- fit_daylight$draws("y_rep", format = "matrix")
y_rep_day_draws <- posterior::as_draws_matrix(fit_daylight$draws("y_rep"))

p_day_hist <- ppc_hist(
  y = y_obs,
  yrep = y_rep_day_mat[seq_len(n_ppc_draws), ]
) +
  labs(title = "Daylight-only — posterior predictive histogram")

p_day_dens <- ppc_dens_overlay(
  y = y_obs[nz_idx],
  yrep = y_rep_day_mat[seq_len(n_ppc_draws), nz_idx]
) +
  labs(title = "Daylight-only — non-zero values only")

p_day_zero <- ppc_stat(y_obs, y_rep_day_draws, stat = prop_zero) +
  labs(title = "Daylight-only — proportion of zeros")

p_day_mean_sd <- ppc_stat_2d(y_obs, y_rep_day_draws, stat = c("mean", "sd")) +
  labs(title = "Daylight-only — mean vs SD")

p_day_hist / p_day_dens
p_day_zero
p_day_mean_sd
```


**Daylight intervention — SCM vs daylight-only**
```{r}
Ey_scm_day <- ey_to_df(fit, "Ey_do_daylight", interv_grid) |>
  mutate(model = "SCM (DAG total effect)")

Ey_bench_day <- ey_to_df(fit_daylight, "Ey_do", interv_grid) |>
  mutate(model = "Daylight-only (direct on Y)")

Ey_day_compare <- bind_rows(Ey_scm_day, Ey_bench_day)

p_day_compare <- ggplot(Ey_day_compare, aes(x = do_value, y = mean, colour = model, fill = model)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 1) +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    title = expression("Daylight intervention:" ~ E[Y * "|" * do(daylight)]),
    subtitle = "SCM propagates through air temp, SST, nutrients, seaweed, and phytoplankton",
    x = "Daylight intervention (standardised)",
    y = "Expected biofouling",
    colour = NULL,
    fill = NULL
  ) +
  theme_bw()

p_day_compare

Ey_day_compare |>
  filter(do_value %in% range(interv_grid)) |>
  select(model, do_value, mean, lo, hi)
```


**PPC comparison — SCM vs daylight-only**
```{r}
ppc_compare <- bind_rows(
  tibble(
    model = "SCM",
    stat = c("zero", "mean", "sd"),
    observed = c(prop_zero(y_obs), mean(y_obs), sd(y_obs)),
    replicated = c(
      mean(apply(y_rep_mat, 1, prop_zero)),
      mean(rowMeans(y_rep_mat)),
      mean(apply(y_rep_mat, 1, sd))
    )
  ),
  tibble(
    model = "Daylight-only",
    stat = c("zero", "mean", "sd"),
    observed = c(prop_zero(y_obs), mean(y_obs), sd(y_obs)),
    replicated = c(
      mean(apply(y_rep_day_mat, 1, prop_zero)),
      mean(rowMeans(y_rep_day_mat)),
      mean(apply(y_rep_day_mat, 1, sd))
    )
  )
)

ppc_compare |>
  mutate(abs_error = abs(observed - replicated)) |>
  pivot_longer(c(observed, replicated), names_to = "type", values_to = "value") |>
  ggplot(aes(x = stat, y = value, fill = type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~model) +
  labs(
    title = "Posterior predictive summary: observed vs replicated",
    x = NULL,
    y = "Value",
    fill = NULL
  ) +
  theme_bw()

p_ppc_bar <- ppc_compare |>
  pivot_longer(c(observed, replicated), names_to = "type", values_to = "value") |>
  ggplot(aes(x = stat, y = value, fill = type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_wrap(~model) +
  labs(
    title = "Posterior predictive summary: observed vs replicated",
    x = NULL,
    y = "Value",
    fill = NULL
  ) +
  theme_bw()

ppc_compare
```


**Save figures**
```{r}
dir.create("figures", showWarnings = FALSE)

ggsave("figures/scm_ppc_hist_dens.png", p_ppc_hist / p_ppc_dens, width = 10, height = 4, dpi = 150)
ggsave("figures/scm_ppc_zero_mean_sd.png", p_ppc_zero / p_ppc_mean_sd, width = 10, height = 4, dpi = 150)
ggsave("figures/scm_intervention_all.png", p_interv_all, width = 11, height = 9, dpi = 150)
ggsave("figures/daylight_intervention_compare.png", p_day_compare, width = 8, height = 5, dpi = 150)
ggsave("figures/ppc_scm_vs_daylight.png", p_ppc_bar, width = 8, height = 4, dpi = 150)

saveRDS(
  list(
    fit_scm = fit,
    fit_daylight = fit_daylight,
    Ey_scm_all = Ey_scm_all,
    Ey_day_compare = Ey_day_compare,
    ppc_compare = ppc_compare
  ),
  "scm_analysis_results.rds"
)
```

